Short paper : Automatic predicate testing in formal certification

### 2016-02-19 ProofsAndTests/paper-TAP/content/1-introduction.tex
paper continued

There are multiple ways for increasing confidence in the software we build. One of them is to formally prove its correctness, using a proof assistant. These proofs assistants enable to write code, logical statements and proof in the same language, and offer the guarantee that every proof will be automatically checked. The appearance of these languages, and the way to write programs in them, varies from one to another. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigm are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logic (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantee tend to be too often consider as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof isn't enough. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

Say we want to implement a formally verified sorting function that works for list of elements of type $T$, where $T$ is ordered by a relation $\leq$.
We can decide to put more or less information inside the type of the sorting function $sort$, in order to make it more or less explicit that this function is precisely a sorting function. We can decide to only use a ``weak" type, like $sort\ :\ \{T:Type\}\ \rightarrow\ List\ T\ \rightarrow\ List\ T$, and to use an external lemma to ensure the correctness of the function. Which property does this function has to respect ? First, the output has to be sorted. So first, we need to define this notion of being sorted. For this, we can define an inductive predicate.

@@

Things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. Then, the function $sort$ could do whatever it wants, its result would be seen as ``sorted". For example, if the bijection condition has been added to the formal specification --and we can't therefore return $[\ ]$--, it would for example be possible to return the exact input, unchanged. Such a function would also unintentionally be seen as ``correct". One could object that the person doing the proof of correctness should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than the next element in the list. The reality is that currently, many effort are going in the direction of proof automation, where the machine builds the proof automatically. Many proof assistants are enhanced regularly with new automations that solves automatically some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005}, a solver for Presburger arithmetic, and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures [ref]. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics.  Thus, the general tendency is precisely to let the machine find the proof on its own, and to only interact with the system when needed. And in the imperative world, the formal B-method also follows the same spirit, with many routine proofs being proven automatically. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user. We can't reasonably demand that the user of the proof assistant should read all of the automatically generated proof, looking for proofs that seem ``too easy" in order to find weaknesses in the specifications.

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification --written in some formal logic-- and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? Too few people realize that the proving activity has created this silent assumption of adequacy between the formalised specification and the original real requirements, usually written in english. Of course, the proving activity has undeniably increased the confidence in the produced software, and the aim of this paper isn't to say otherwise. Many successful applications of these formal methods, such as a formally verified C compiler \cite{Leroy09} have shown the value of developing a machined-checked proof of correctness. Instead, the primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
\item Show the usual and naive approaches to the problem of a badly specified function which leads to a proof that does not convey the expected guarantee (section 2)
\item Show the distinction between structural properties, and the properties on the content, and how to identify what needs to be tested for the former category (section 3).
\item Show a way to conduct these tests in the proof assistant itself, by automatically generating terms, and we present a technique to automate these tests as far as possible (section 4)
\end{itemize}

>>>>

There are multiple ways for increasing confidence in the software we build. One of them is to formally prove its correctness, using a proof assistant. These proofs assistants enable to write code, logical statements and proof in the same language, and offer the guarantee that every proof will be automatically checked. The appearance of these languages, and the way to write programs in them, varies from one to another. Many of them are functional programming languages, like Coq, Idris and Agda, and others, like the B-Toolkit belong to the imperative paradigm. These different paradigm are internally supported by different logic. Systems like Coq, Idris and Agda are based on various higher order logic (CoC, a variant of ML and LUO respectively) and are realisations of the Curry-Howard correspondence, while the formal B method is based on Hoare logic. These different foundations lead to different philosophies and different ways to implement and verify a software, but all of them greatly increase the confidence of the produced software. However, these guarantee tend to be too often consider as perfect, when they are in fact far from it. Knuth was --certainly ironically-- saying ``Beware of bugs in the above code; I have only proved it correct, not tried it". The reality is precisely that a proof isn't enough\footnote{That's without taking in account that when the proof is only done on the paper (and not machine-checked), the proof can be wrong. The paper never complains about a false proof}. When we prove the correctness of a function, we only gain the guarantee expressed by the proven lemma, and nothing more.

Say we want to implement a formally verified sorting function that works for list of elements of type $T$, where $T$ is ordered by a relation $\leq$.
We can decide to put more or less information inside the type of the sorting function $sort$, in order to make it more or less explicit that this function is precisely a sorting function. We can decide to only use a ``weak" type, like $sort\ :\ \{T:Type\}\ \rightarrow\ List\ T\ \rightarrow\ List\ T$, and to use an external lemma to ensure the correctness of the function. Which property does this function has to respect? First, the output has to be sorted. So first, we need to define this notion of being sorted. For this, we can define an inductive predicate.

@@

Things more nasty can happen. Imagine that the third constructor of $isSorted$, called $consSorted$ would have been written with a typo, and that the condition $(h1 \leq h2) $ would have been incorrectly written as $(h1 \leq h1)$. Then, the function $sort$ could do whatever it wants, its result would be seen as ``sorted". Said differently, any list is seen as ``sorted" according to this predicate, just because of this single typo. For example, if the bijection condition has been added to the formal specification --and therefore the function can no longer return $[\ ]$ all the time--, it would for example be possible to return the exact input, unchanged. Such a function would also unintentionally be seen as ``correct". One could object that the person doing the proof of correctness should realize that the proof is being done too easily, without having to use the essential property that the output is being built such as any element in the list is always lower or equal than the next element in the list. The reality is that currently, many effort are going in the direction of proof automation, where the machine builds the proof automatically. Many proof assistants are enhanced regularly with new automations that solves automatically some kind of goals. For example, Coq has already a Ring prover~\cite{coq2005}, a solver for Presburger arithmetic, and many others automations, and Idris has been recently equipped with a hierarchy of provers for algebraics structures [ref]. There are even extensions to languages, such as Ltac~\cite{DelahayeLTac} and Mtac~\cite{Ziliani13} that aim to help the automation of tactics.  Thus, the general tendency is precisely to let the machine find the proof on its own, and to only interact with the system when needed. And in the imperative world, the formal B-method also follows the same spirit, with many routine proofs being proven automatically. The problem is that the machine is never going to find a proof ``too easy", and will never report that something seems weird with the specification given by the user. We can't reasonably demand that the user of the proof assistant should read all of the automatically generated proof, looking for proofs that seem ``too easy" in order to find weaknesses in the specifications.

Thus, if we want to trust the proven software, we're now forced to believe that there is adequacy between the formal specification --written in some formal logic-- and the informal requirements. A switch has occurred. We used to have to trust code, but we now have to trust logical statements and predicates. But when the specification is too often as complicated as the code, why should we blindly believe in it, when we've first refused to blindly trust the code? Too few people realize that the proving activity has created this silent assumption of adequacy between the formalised specification and the original real requirements, usually written in english. Of course, the proving activity has undeniably increased the confidence in the produced software, and the aim of this paper isn't to say otherwise. Many successful applications of these formal methods, such as a formally verified C compiler \cite{Leroy09} have shown the value of developing a machined-checked proof of correctness. Instead, the primary aim of this paper is to raise awareness of the adequacy concern, and to see how heterogeneous approaches, that mix both proofs and tests, can help to go a step forward in the certification process, in the context of proof assistants based on type theory.
More precisely, we :
\begin{itemize}
	\item Show the usual and naive approaches to the problem of a badly specified function which leads to a proof that does not convey the expected guarantee (section~\ref{sect:naiveApproaches})
	\item Show the distinction between structural properties, and the properties on the content, and how to identify what needs to be tested for the former category (section~\ref{sect:differentKindOfProperties}).
	\item Present a way to conduct these tests in the proof assistant itself, by automatically generating terms, and we present a technique to automate these tests as far as possible (section~\ref{sect:testingInside})
\end{itemize}

### 2016-02-19 ProofsAndTests/paper-TAP/content/2-specCanGoWrong.tex
paper continued

In order to gain confidence in the formal specifications we write, an easy an naive approach is to test the predicate itself. That consists of defining a few terms, usually by hand, for which we know if the predicate should hold or not, and to prove that the predicate effectively holds when it should, and that it doesn't when it shouldn't.

@@

\begin{lstlisting}
isSorted_test1 : isSorted natIsOrdered [3, 5, 7]
isSorted_test1 = 
  let p1 : (3 <= 5) = tryDec (lowerEqDec nat_partialOrder 3 5) in
  let p2 : (5 <= 7) = tryDec (lowerEqDec nat_partialOrder 5 7) in
    ConsSorted 3 5 [7] (ConsSorted 5 7 [] (SingletonIsSorted _ 7) p2) p1
\end{lstlisting}

It is important to realize that this kind of tests is very particular in the sense that is is a \emph{test by proof} : we show that the predicate holds by doing the proof on a specific value, here $[3,\ 5,\ 7]$.

We could go a step forward, and remove the need of doing these specific proofs by hand, because in this case, the predicate isSorted can be decided : there exists an algorithm that decided if a list $l$ is sorted --according to isSorted-- and that produces a proof of $(isSorted\ l)$ if appropriate, or a proof of $(not\ (isSorted\ l))$ otherwise.

@@

- Other usual approach to mention : defining many of them and proving their equivalence,
[quarter of a page]

>>>>

\label{sect:naiveApproaches}

When confronted to this problem of adequacy between the intuitive notion and the formalised one, a first possibility that usually requires quite a lot of effort is to formalise the notion multiple times, with different predicates, and to prove that they are equivalent.
With our example, that means that we need to find another formalisation $isSorted'$ of being sorted, and to prove the following lemma.
$pred\_equiv : \forall\ (T:Type)\ (l:List\ T), (isSorted\ l) \leftrightarrow (isSorted'\ l)$.
This approach aims at increasing the confidence in our formal definitions by assuming that if we've managed to define multiple times the same notion, then we've surely succeeded to define the notion we wanted. The biggest problem with this approach is to be able to find some alternative formalisations that are sufficiently different from the original one. Obviously, if the new normalisations are too similar to the original one (and in the worst case the new ones are just syntactical variants of the first one), then we won't gain any guarantee. The ideal is to be able to capture the same notion by using very different point of view, but this task can be incredibly difficult, and the more different they are, the harder the equivalence proofs will be.

\

In order to gain confidence in the formal specifications we write, another traditional approach is to test the predicate on some values. That consists of defining a few terms, usually by hand, for which we know if the predicate should hold or not, and to prove that the predicate effectively holds when it should, and that it doesn't when it shouldn't.

@@

\begin{lstlisting}
isSorted_test1 : isSorted natIsOrdered [3, 5, 7]
isSorted_test1 = 
  let p1 : (3 <= 5) = tryDec (lowerEqDec natIsOrdered 3 5) in
  let p2 : (5 <= 7) = tryDec (lowerEqDec natIsOrdered 5 7) in
    ConsSorted 3 5 [7] (ConsSorted 5 7 [] (SingletonIsSorted _ 7) p2) p1
\end{lstlisting}

It is important to realize that this kind of tests is very particular in the sense that it is a \emph{test done by proof} : we show that the predicate holds on some specific value, here $[3,\ 5,\ 7]$, by doing the proof.

We could go a step forward by removing the need of doing these specific proofs by hand, because in this case, the predicate isSorted can be decided : there exists an algorithm that decided if a list $l$ is sorted --according to isSorted-- and that produces a proof of $(isSorted\ l)$ if appropriate, or a proof of $(not\ (isSorted\ l))$ otherwise.

@@

Now, in order to do \emph{tests by proof} on the predicate $isSorted$, we can simply run the above decision procedure.

\begin{lstlisting}
isSorted_test1' : Dec (isSorted natIsOrdered [3,5,7])
isSorted_test1' = decideIsSorted natIsOrdered [3,5,7] 
\end{lstlisting}

And if we ask to evaluate $isSorted\_test1'$, the system will answer $Yes$ and the automatically generated proof of $isSorted\ [3,\ 5,\ 7]$, which is what we want, and we therefore know that the predicate has passed this test.

With this technique, we can run semi-automatically a few tests on the predicate $isSorted$. It is semi-automatic in the sense that we still have to define by hand some terms that we know sorted and unsorted --according to our intention-- but we can let the machine produce the proof that the predicate holds or not on these specific values. This isn't bad --and this is in fact all that is done usually, when it is actually done-- but we would like to have a stronger guarantee, and not only that the predicate will coincide with our intuitive notion on a couple of tested terms. We would like to make sure that our intuitive notion and the formalised one coincide for every term.
